Almost orthogonal vectors

### 2014-12-09 project3/paper/paper.tex
Changes to the paper -- added code for algorithm, and also corrected error in proof.

% Special symbols, etc.
\usepackage{amssymb,amsbsy,latexsym,ytableau}
\usepackage{amsmath}
\usepackage{graphics, subfigure, float} 
\usepackage{cancel}
%\usepackage{todonotes}
% Encoding settings
\usepackage[latin1]{inputenc}
\usepackage[american]{babel}
\usepackage[T1]{fontenc} 
\usepackage{tikz}

@@

\usepackage{todonotes}

@@

We shall also see that, for $N=n+1$, the minimum value of $\epsilon$ will be upper bounded by
\[\frac{1}{n^2},\] 
and this configuration is produced when the $n+1$ vectors point to the vertices of the $n$-simplex.

@@

It's clear that when $N=3$, the value of $\epsilon$ must be greater than $0$, since there is no way three dimensionality-$2$ vectors can be orthogonal to each other.

In addition, we see that for $N=3$, the quantity $\epsilon$ for three unit-length vectors $v_1, v_2, v_3 \in \mathbb{R}^2$ is given by $$\max \{ |v_1 \cdot v_2|^2, |v_1 \cdot v_3|^2, |v_2 \cdot v_3|^2 \}$$

Since we don't care about the sign of the dot product between any two vectors, if we define the unit vectors $v_4, v_5, v_6 \in \mathbb{R}^2$ to be $-v_1, -v_2, -v_3$ respectively, then we see that $\epsilon$ can be equivalently expressed as $$\max_{i \neq j} \{|v_i \cdot v_j |^2 \}$$
where $i, j \in \{1,2,\ldots,6\}$ -- the above result holds because $$|v \cdot v'| = |(-v) \cdot v'| = |(-v) \cdot (-v')| = |v \cdot (-v')|$$ for some arbitrary vector $v'$. Furthermore, we see that there now exists an axis of symmetry -- one of the diagonals passing through either $v_1$ and $v_4 = -v_1$ or $v_2$ and $v_5 = -v_2$ or $v_3$ and $v_6 = -v_3$ -- so we don't need to look at the entire unit circle anymore, and instead can focus on the four vectors contained in a semi-circle resting on one of these three diameters.

Without loss of generality, let $x_1 = [1, 0]^T$; then since $x_4$ is $-x_1$, we see that $x_4 = [-1, 0]^T$. Also, without loss of generality let us assume that $x_2$ and $x_3$ are above the $x$-axis, and that $x_2$ is to the right of $x_3$.

@@

        \node at (4, -0.3) {\text{$x_1 = [1,0]$}};
        \node at (2.8, -3.1) {\text{$x_2 = [\cos \pi/3,\sin \pi/3]$}};
        \node at (-2.8, -3.1) {\text{$x_3 = [\cos 2\pi/3,\sin 2\pi/3]$}};
    \end{tikzpicture}
    \caption{A configuration of three unit vectors that produce the optimum $\epsilon$ for $n=2, N=3$}
\end{figure}

Let $\theta_1$ be the angle between $x_1$ and $x_2$, $\theta_2$ be the angle between $x_2$ and $x_3$ and $\theta_3$ be the angle between $x_3$ and $x_4$. Now, we see that $\theta_1 + \theta_2 + \theta_3 = \pi$ (since vectors $v_1$ and $v_4 = -v_1$ are anti-parallel) and that $$\epsilon = \max_{i \in \{1,2,3\}} \cos ^2 \theta_i$$, since $|x_i \cdot x_j| = \cos \theta$ where $\theta$ is the angle between the two vectors $x_i$ and $x_j$.

Before proceeding, we state and prove the following lemma.
\begin{lemma}
Consider $n$ angles $\theta_1, \theta_2, \ldots, \theta_n \in [0, \pi]$ s.t. $\theta_1 + \theta_2 + \ldots + \theta_n = \pi$. Then $\epsilon = \max_i \cos^2 \theta_i$ must equal $\cos^2 \theta_j$ where $j = \text{argmin }\theta_i$.
\end{lemma}

Our arguments will hinge on the fact that for all $\theta \in [0, \pi/2]$, the function $\cos \theta$ is decreasing -- this is easy to see from Figure $2$. Without loss of generality, let us assume that $\theta_1 \geq \theta_2 \geq \ldots \geq\theta_n$. Hence our theorem statement is equivalent to proving that $\epsilon$ is equal to $\cos^2 \theta_n$.

We split our proof into two cases,
\begin{itemize}
\item $\theta_1, \theta_2, \ldots, \theta_n \in [0, \pi/2]$: In this case, it is easy to see that $\epsilon = \cos^2 \theta_n$ from the fact that $\cos \theta$ is a decreasing function in $\theta$ if $\theta \in [0, \pi/2]$.

In addition, we see that for all $i \in \{2, 3, \ldots, n-1\}$, $\cos^2 \theta_i < \cos^2 \theta_n$, from which we can conclude that in this case too, $\epsilon$ is equal to $\cos^2 \theta_n$.

@@

@@

Again without loss of generality, we can assume that the vectors $x_1, x_2, \ldots, x_N$ are on or above the $x$-axis, and that $x_1 = [1,0]^T$ and that $x_{N+1} = [-1,0]^T$ -- if any vector $x_i$ were not above the $x$-axis, then we could just consider $-x_i$ instead.

Let us define $\theta_i$ as the angle between the vectors $x_i$ and $x_{i+1}$. Then, we quickly see that $\theta_1 + \theta_2 + \ldots + \theta_N = \pi$. If $j = \text{argmin } \theta_i$, $\epsilon = \cos^2 \theta_j$ by the above Lemma.

Since $\theta_ 1 + \theta_2 + \ldots + \theta_N \geq N \cdot \theta_j \Rightarrow \theta_j \leq \pi/N$, we conclude that the optimum $\epsilon$ value is in fact equal to $\cos^2 \pi/N$ (again making use of the fact that the cosine function is decreasing).
\end{proof}

@@

\section{The Parallelogram Law}

@@

\section{$N = n+1$}
Let us consider the situation in which we try to minimize $\epsilon$ for $n+1$ vectors in an $n$ dimensional space. First we introduce the definition of a simplex, and then we will proceed to characterize $\epsilon$.

@@

\section{Simulated annealing and associated conjectures}
Since visualizing in spaces of dimensionality greater than $3$ is extremely hard, we used an annealing approach to compute near-optimal epsilon values for different $(n, N)$ pairs. We describe this method in greater detail below.

Our general intuition tells us that perturbing a set of nearly optimal vectors slightly could give us a new set of vectors that produce even a smaller epsilon value. Given this, we start off with a set of random vectors, and try to move these vectors in order to produce smaller epsilon values.

At every iteration, we maintain the set of vectors (called cur\_vectors) that have produced the optimal epsilon value so far. Then we compute a set new\_vectors defined as follows,
$$\text{new\_vector} = \text{old\_vector} + (1/i \cdot \text{random\_vector})$$

We update cur\_vectors to the set new\_vectors if the epsilon value produced by new\_vectors is less than the epsilon value produced by cur\_vectors. $i$ is incremented whenever cur\_vectors is updated -- having the $1/i$ factor in the above expression ensures that as we move closer to an optimal value, we perturb the set of vectors by a smaller amount.

@@

To make this more concrete, consider Table $1$.
\begin{table}
   \centering
    \begin{tabular}{ | c | c | c | c |}
    \hline
      & $n=2$ & $n=3$ & $n=4$ \\ \hline
    $N=2$ & $0.0$ & $-$ & $-$ \\ \hline
    $N=3$ & $0.25$ & $0.0$ & $-$ \\ \hline
    $N=4$ & $0.5$ & $0.11$ & $0.0$ \\ \hline
    $N=5$ & $0.66$ & $0.21$ & $0.07$ \\ \hline
    $N=6$ & $0.75$ & $0.22$ & $0.14$ \\ \hline
    $N=7$ & $0.82$ & $0.36$ & $0.18$ \\ \hline
    $N=8$ &  $0.86$ & $0.44$ & $0.24$ \\ \hline
    $N=9$ &  $0.89$ & $0.49$ & $0.26$ \\ \hline
    $N=10$ & $0.91$ & $0.51$ & $0.32$ \\ \hline
    $N=11$ & $0.93$  & $0.55$ & $0.35$ \\ \hline
    $N=12$ & $0.94$ & $0.64$ & $0.41$ \\
    \hline
    \end{tabular}
    \caption {Minimum values of $\epsilon$ for different values of $(n, N)$, as computed by our annealing method}
\end{table}

The accuracy of our method can be noted by looking at the computed values for $n=2$ -- the computed values for different values of $N$ are very close to the expected value of $\cos^2 \pi/N$.
\section{Applications to sphere packing}
We should be able to prove something about the radius of spheres you can pack in the case of $N=n+1$, taking as intuition square and hexagonal tilings of the plane and cubic and tetrahedral tilings of 3-space.

>>>>

% Special symbols, etc.
\usepackage{amssymb,amsbsy,latexsym,ytableau}
\usepackage{amsmath}
\usepackage{graphics, subfigure, float} 
\usepackage{cancel}
%\usepackage{todonotes}
% Encoding settings
\usepackage[latin1]{inputenc}
\usepackage[american]{babel}
\usepackage[T1]{fontenc} 
\usepackage{tikz}
\usepackage[scaled]{beramono}
\usepackage{listings}

\lstset{
  language=Python,
  showstringspaces=false,
  formfeed=\newpage,
  tabsize=4,
  commentstyle=\itshape,
  basicstyle=\ttfamily,
  morekeywords={while, if}
  captionpos=b,
}

@@

\renewcommand{\lstlistingname}{Algorithm}

\usepackage{todonotes}
\usepackage{caption}
\DeclareCaptionFormat{myformat}{#1#2#3\hrulefill}
\captionsetup{format=myformat}
\captionsetup[lstlisting]{position=bottom,format=myformat}

@@

We shall also see that, for $N=n+1$, the minimum value of $\epsilon$ is upper bounded by
\[\frac{1}{n^2},\] 
and this configuration is produced when the $n+1$ vectors point to the vertices of the $n$-simplex.

@@

It's clear that when $N=3$, the value of $\epsilon$ must be greater than $0$, since there is no way three dimensionality-$2$ vectors can be orthogonal to each other. We make this more concrete in the following theorem.

\begin{theorem}
For $n=2$ and $N=3$, the minimum possible $\epsilon$ is equal to $1/4$.
\end{theorem}

Before proving the above theorem, we state and prove the following lemma.

\begin{lemma}
Consider $n$ angles $\theta_1, \theta_2, \ldots, \theta_n \in [0, \pi]$ s.t. $\theta_1 + \theta_2 + \ldots + \theta_n = \pi$. Then $\alpha = \max_i \cos^2 \theta_i$ must equal $\cos^2 \theta_j$ where $j = \text{argmin }\theta_i$.
\end{lemma}

Our arguments will hinge on the fact that for all $\theta \in [0, \pi/2]$, the function $\cos \theta$ is decreasing -- this is easy to see from Figure $1$. Without loss of generality, let us assume that $\theta_1 \geq \theta_2 \geq \ldots \geq\theta_n$. Hence our theorem statement is equivalent to proving that $\alpha$ is equal to $\cos^2 \theta_n$.

We split our proof into two cases,
\begin{itemize}
\item $\theta_1, \theta_2, \ldots, \theta_n \in [0, \pi/2]$: In this case, it is easy to see that $\alpha = \cos^2 \theta_n$ from the fact that $\cos \theta$ is a decreasing function in $\theta$ if $\theta \in [0, \pi/2]$.

In addition, we see that for all $i \in \{2, 3, \ldots, n-1\}$, $\cos^2 \theta_i < \cos^2 \theta_n$, from which we can conclude that in this case too, $\alpha$ is equal to $\cos^2 \theta_n$.

Given this lemma, we now prove the above theorem.

\begin{proof}
Observe that for $N=3$, the quantity $\epsilon$ for three unit-length vectors $v_1, v_2, v_3 \in \mathbb{R}^2$ is given by $$\max \{ |v_1 \cdot v_2|^2, |v_1 \cdot v_3|^2, |v_2 \cdot v_3|^2 \}$$

Since we don't care about the sign of the dot products between any two vectors (we're looking at the squares of the dot products in the above equation), and since $$|v \cdot v'| = |v \cdot (-v')|$$ for some arbitrary vector $v'$, we see that it's easy to transform the three vectors $v_1$, $v_2$ and $v_3$ so that they all lie in the same semi-circle -- to accomplish this, at most one vector needs to be reflected about the origin.

Without loss of generality, let $v_1 = [1, 0]^T$. Also, without loss of generality let us assume that $v_2$ and $v_3$ are above the $x$-axis, and that $v_1$, $v_2$ and $v_3$ are in anti-clockwise order.

@@

        \node at (4, -0.3) {\text{$v_1 = [1,0]$}};
        \node at (2.8, -3.1) {\text{$v_2 = [\cos \pi/3,\sin \pi/3]$}};
        \node at (-2.8, -3.1) {\text{$v_3 = [\cos 2\pi/3,\sin 2\pi/3]$}};
    \end{tikzpicture}
    \caption{A configuration of three unit vectors that produce the optimum $\epsilon$ for $n=2, N=3$}
\end{figure}

Let $\theta_1$ be the angle between $v_1$ and $v_2$ and $\theta_2$ be the angle between $v_2$ and $v_3$. Let us now define $\theta_3$ such that $\theta_1 + \theta_2 + \theta_3 = \pi$. Observe that $|v_1 \cdot v_3|^2 = \cos^2(\theta_1 + \theta_2) = \cos^2 (\pi - \theta_3) = \cos^2 \theta_3$, hence we can conclude that

$$\epsilon = \max_{i \in \{1,2,3\}} \cos ^2 \theta_i$$

@@

\end{proof}

@@

Again without loss of generality, we can assume that the vectors $v_1, v_2, \ldots, v_N$ are on or above the $x$-axis, and that $v_1 = [1,0]^T$ -- if any vector $v_i$ were not above the $x$-axis, then we could just consider $-v_i$ (the reflection of $v_i$ about the $x$-axis) instead.

Let us define $\theta_i$ as the angle between the vectors $v_i$ and $v_{i+1}$ for $i \in \{1, 2, \ldots, N-1\}$, and let $\theta_N$ be the angle such that $\theta_1 + \theta_2 + \ldots + \theta_N = \pi$.

Observe that as before, we are interested in maximizing the square of the cosine of the angle between any two vectors in $v_1, v_2, \ldots, v_N$. Note that here, the angle between any two vectors in $v_j$ and $v_k$ in $v_1, v_2, \ldots, v_N$ can be expressed as $\cos^2 (\sum_{i=j} ^{k-1} \theta_i)$. Note that, however, if $\sum_{i=j}^{k-1} \theta_i \leq \pi/2$, then $\cos^2 \theta_j \geq \cos^2(\sum_{i=j}^{k-1} \theta_i)$ and that if $\sum_{i=j}^{k-1} \theta_i > \pi/2$, then $\cos^2 (\sum_{i=j}^{k-1} \theta_i) = \cos^2 (\pi - \sum_{i=j}^{k-1} \theta_i)$ which is less than $\cos^2 \theta_i$ for any $i$ not in $\{j, j+1, \ldots, k-1\}$.

From this we can conclude that $\epsilon$ is equal to $\max_i \cos^2 \theta_i$. Then, if $j = \text{argmin } \theta_i$, then we see that $\epsilon = \cos^2 \theta_j$ by the above Lemma.

Since $\theta_ 1 + \theta_2 + \ldots + \theta_N \geq N \cdot \theta_j \Rightarrow \theta_j \leq \pi/N$, we conclude that the optimum $\epsilon$ value is in fact equal to $\cos^2 \pi/N$ (again making use of the fact that the cosine function is decreasing between $0$ and $\pi/2$).
\end{proof}

@@

\section{$N = n+1$}

We now look to calculate the optimum value of $\epsilon$ for the specific case of $N = n+1$. Before proving an upper bound on what the optimum value of $\epsilon$ can be, we look at the parallelogram law.

\subsection{The Parallelogram Law}

@@

\subsection{Upper bound on minimum possible $\epsilon$}
Let us consider the situation in which we try to minimize $\epsilon$ for $n+1$ vectors in an $n$ dimensional space. First we introduce the definition of a simplex, and then we will proceed to characterize $\epsilon$.

@@

\section{Simulated annealing and associated conjectures}

Since visualizing in spaces of dimensionality greater than $3$ is extremely hard, we used an annealing approach to compute near-optimal epsilon values for different $(n, N)$ pairs. We describe this method in greater detail below.

Our general intuition tells us that perturbing a set of nearly optimal vectors slightly could give us a new set of vectors that produce even a smaller epsilon value. Given this, we start off with a set of random vectors, and try to move these vectors towards smaller epsilon values.

At every iteration, we try adding a vector whose norm becomes smaller with every passing iteration, to the already computed optimal vector.

\begin{lstlisting}
def get_min_epsilon_annealing(n, N, num_iter, start_vectors):
    min_epsilon = 1.0  # Start off with the worst possible epsilon
    current_vectors = start_vectors
    i = 0
    while (i < num_iter):
        temperature = 1.0 / float(i + 1)

        # Get perturbing vectors
        random_vectors = get_random_vectors(n, N)

        # Now perturb current_vectors
        new_vectors = list()
        for j in xrange(N):
            new_vector = list()
            for k in xrange(n):
                new_vector.append(
                    current_vectors[j][k] + (
                    	temperature * random_vectors[j][k]))
            new_vectors.append(new_vector)
        normalize_vectors(new_vectors)

        # Accept change only if new_vectors produces a better epsilon
        epsilon = compute_epsilon(new_vectors)
        if epsilon < min_epsilon:
            min_epsilon = epsilon
            current_vectors = new_vectors
            i += 1

    return min_epsilon, current_vectors
\end{lstlisting}
\captionof{lstlisting}{The annealing algorithm used to generate optimum epsilon values}

To make this more precise, we present Python code for the algorithm above.

@@

To make this more concrete, consider Table $1$.

The accuracy of our method can be noted by looking at the computed values for $n=2$ -- the computed values for different values of $N$ are very close to the expected value of $\cos^2 \pi/N$, that we already proved in Section $2$, which makes us confident about the accuracy of this numeric method.
\section{Applications to sphere packing}
We should be able to prove something about the radius of spheres you can pack in the case of $N=n+1$, taking as intuition square and hexagonal tilings of the plane and cubic and tetrahedral tilings of 3-space.

\begin{table}
   \centering
    \begin{tabular}{ | c | c | c | c |}
    \hline
      & $n=2$ & $n=3$ & $n=4$ \\ \hline
    $N=2$ & $0.0$ & $-$ & $-$ \\ \hline
    $N=3$ & $0.25$ & $0.0$ & $-$ \\ \hline
    $N=4$ & $0.5$ & $0.11$ & $0.0$ \\ \hline
    $N=5$ & $0.66$ & $0.21$ & $0.07$ \\ \hline
    $N=6$ & $0.75$ & $0.22$ & $0.14$ \\ \hline
    $N=7$ & $0.82$ & $0.36$ & $0.18$ \\ \hline
    $N=8$ &  $0.86$ & $0.44$ & $0.24$ \\ \hline
    $N=9$ &  $0.89$ & $0.49$ & $0.26$ \\ \hline
    $N=10$ & $0.91$ & $0.51$ & $0.32$ \\ \hline
    $N=11$ & $0.93$  & $0.55$ & $0.35$ \\ \hline
    $N=12$ & $0.94$ & $0.64$ & $0.41$ \\
    \hline
    \end{tabular}
    \caption {Minimum values of $\epsilon$ for different values of $(n, N)$, as computed by our annealing method}
\end{table}
